A Classical Density-Functional Theory for Describing Water Interfaces

### 2013-06-27 papers/water-saft/paper.tex
Started work on new water paper.  Started using python to generate some figures. Tested everything on node10.

@@

\subsection{Statistical Associating Fluid Theory}

@@

The SAFT Helmholtz free energy is composed of five terms:
\begin{equation} \label{eq:SAFT-free-energy}
  F = F_\textit{id} + F_\textit{hs} + F_\textit{disp} +
  F_\textit{assoc} + F_\textit{chain},
\end{equation}
where the first three terms---ideal gas, hard-sphere repulsion and
dispersion---encompass the \emph{monomer} contribution
to the free energy, the fourth is the \emph{association} free energy,
describing hydrogen bonds, and the final term is the chain formation
energy for fluids that are chain polymers. In our model, we use a hard
sphere reference fluid which renders $F_\textit{chain}$ irrelivant for
our functional.

In a recent paper \cite{hughes2013classical}, Hughes \emph{et al}
constructed a classical density functional theory for water which used
Clark \emph{et al}'s \cite{clark2006developing} five fitting
paramenters for an optimal cDFT for water.  In addition, they
introduced a parameter for the length scale over which the dispersion
interaction is correlated

@@

Schulte \emph{et al}\cite{schulte2012using} derived expressions for
the correlation fucntoin at contact based on the White Bear version of
FMT. They found their $g_{\sigma}^A$ to agree with Monte Carlo
simulation than Yu and Wu's which we was used in Hughes \emph{et al}'s
paper.  We use Hughs \emph{et al}'s functional and replace the
correlation with Schulte \emph{et al}'s, which we will call
$g_{\sigma}$. \fixme{expand this}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figs/single-rod-X-plot}
\end{center}
\caption{ The average number of hydrogen bonding per molecule at
  distances from the center of a hydrophobic rod for various
  radii. The solid lines are the results from using the older
  $g_{\sigma}$, while the dotted lines are from Schulte \emph{et al}'s
  asymmetric $g_{\sigma}$.}
\label{fig:single-rod-X}
\end{figure}

@@

\newcommand\etadisp{\ensuremath{\eta_\textit{d}}}
\newcommand\epsilondisp{\ensuremath{\epsilon_\textit{d}}}
\newcommand\epsilonassoc{\ensuremath{\epsilon_\textit{a}}}
\newcommand\kappaassoc{\ensuremath{\kappa_\textit{a}}}
\newcommand\lambdadisp{\ensuremath{\lambda_\textit{d}}}
\newcommand\lscale{\ensuremath{s_d}}
\subsection{Dispersion free energy}
The dispersion free energy includes the van der Waals attraction and
any orientation-independent interactions. We use a dispersion term
based on the SAFT-VR approach\cite{gil-villegas-1997-SAFT-VR}, which
has two free parameters (taken from Clark~\emph{et
  al}\cite{clark2006developing}): an interaction energy $\epsilondisp$
and a length scale $\lambdadisp R$.

@@

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figs/density-single-rod}
\end{center}
\caption{ Density profiles for single rods of different radii. The dotted line
represents the saturated liquid density and the points represent the
expected contact density derived from the contact value theorem and
calculated free energy data.}
\label{fig:density-single-rod}
\end{figure}

>>>>

\newcommand\etadisp{\ensuremath{\eta_\textit{d}}}
\newcommand\epsilondisp{\ensuremath{\epsilon_\textit{d}}}
\newcommand\epsilonassoc{\ensuremath{\epsilon_\textit{a}}}
\newcommand\kappaassoc{\ensuremath{\kappa_\textit{a}}}
\newcommand\lambdadisp{\ensuremath{\lambda_\textit{d}}}
\newcommand\lscale{\ensuremath{s_d}}

@@

@@

\fixme{Talk about Wertheim's TPT}

In a recent paper \cite{hughes2013classical}, Hughes \emph{et al}
constructed a classical density functional theory for water which uses
SAFT as a basis for their free energy functional.  Hughes employed Clark
\emph{et al}'s \cite{clark2006developing} five fitting parameters for
an optimal cDFT for water and, in addition, introduces a parameter for
the length scale over which the dispersion interaction is correlated.
Clark found that the four association site model works best to
describe water, which is physically accurate since each water molecule
has four hydrogen bonding sites.

Schulte \emph{et al}\cite{schulte2012using} derived two expressions
for the correlation function at contact for inhomogeneous fluids based
on the White Bear version of fundamental measure theory (FMT). One
expression was labeled as ``symmetric'' and found the average number
of spheres touching a point $\rr$. The other expression, labeled
``assymetric,'' sets $\rr$ at the center of a sphere, and finds the
average spheres touching the solvent accessable surface.  They found
the asymmetric correlation function, $g_{\sigma}^A$, to agree
closely with Monte Carlo simulation.

@@

We modify Hughes \emph{et al}'s method of creating a classical DFT
based on SAFT using Clark's four site model with five fitting
parameters and the added fitting parameter for the correlation
distance of the dispersion interaction.  Hughe's association free
energy term is modified to use Schulte's $g_{\sigma}^A$ rather than Yu
and Wu's.  The association free energy for our four-site model has the
form \fixme{I copy/pasted and modified these equations, but they still
  require a number of other equations to come before.  Maybe only show
  the $g_\sigma^A$?}
\begin{align}
  F_\text{assoc}[n] &= 4 k_BT \int n_0(\xx) \zeta(\xx)
  \left(\ln X(\xx) - \frac{X(\xx)}{2} + \frac12\right) d\xx,
\end{align}
where the factor of $4$ comes from the four association sites per
molecule, the functional $X$ is the fraction of association sites
\emph{not} hydrogen-bonded, and $\zeta(\xx)$ is a dimensionless
measure of the density inhomogeneity.
\begin{align}
  \zeta(\xx) &= 1 - \frac{\mathbf{n}_{V2}\cdot\mathbf{n}_{V2}}{n_2^2}.
\end{align}
%
The fraction $X$ is determined by the quadratic equation
\begin{align}
  X(\xx) &= \frac{\sqrt{1 + 8n_0(\xx)\zeta(\xx)
      \Delta(\xx)} - 1}
  {4 n_0(\xx)\zeta(\xx)
    \Delta(\xx)}, \label{eq:X}
\end{align}a
where the functional $\Delta$ is a measure of hydrogen-bonding
probability, given by
\begin{align}
  \Delta(\xx) &= \kappaassoc g^\textit{SW}_\sigma(\xx)
  \left(e^{-\beta\epsilonassoc} - 1\right) \\
  g^\textit{SW}_\sigma(\xx) &= g^\textit{A}_\sigma(\xx) +
  \frac{1}{4}\beta\left(\frac{\partial a_1}{\partial \etadisp(\xx)} -
  \frac{\lambdadisp}{3 \etadisp}\frac{\partial a_1}{\partial \lambdadisp}\right)\label{eq:gSW},
\end{align}
% This is equation 77 of gil-villegas-1997-SAFT-VR
where $g^\textit{SW}_\sigma$ is the correlation function evaluated at
contact for a hard-sphere fluid with a square-well dispersion
potential, and $a_1$ and $a_2$ are the two terms in the dispersion
free energy.  The correlation function $g^\textit{SW}_\sigma$ is
written as a perturbative correction to the hard-sphere fluid
correlation function $g^\textit{A}_\sigma$, for which we use the
functional of Schulte \emph{et al}\cite{schulte2012using}:
\begin{align}
  g_\sigma^{A}(\rr) &= \frac{1}{n(\rr)n_A(\rr)}\frac{1}{k_B T 4 \pi
    (2r)^2}\frac{\delta F_{hs}}{\delta R(\rr)} 
\end{align}
where the $n_A(\rr)$ is the density of spheres that are touching a
sphere that is located at point $\rr$.\fixme{The equation for
$g_\sigma^A$ requires finidng convolutions of local derivative of
the free energy, aking this formulation computationally more
expensive than the free energy itself.}

We used Clark's five empirical parameters, and fit the calculated
surface tension to experimental surface tension at ambient
conditions.  This fit was found by choosing the value of the
$\lscale$, which we found to be 0.454 for this new functional.

@@

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figs/density-compare}
\end{center}
\caption{ Density profiles for single rods of different
  radii from old and new functionals.}
\label{fig:density-single-rod}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figs/single-rod-X-plot}
\end{center}
\caption{ The average number of hydrogen bonds per molecule at
  distances from the center of hydrophobic rods of various
  radii. The solid lines are the results from using the assymetric
  $g_{\sigma}^A$, while the dotted lines are from Yu and Wu's $g_{\sigma}$.}
\label{fig:single-rod-X}
\end{figure}

\section{Hydrogen bonding near a hydrophobic rod}

We use X from eq. \ref{eq:X} above to find the average number of
hydrogen bonds per molecule as a function of distance for different
radius hydrophobic rods.  Figure \ref{fig:single-rod-X} compares our
new functional with Hughes's functional, \fixme{Physical description
  for why it looks that way}

@@

\subsection{Dispersion free energy}
The dispersion free energy includes the van der Waals attraction and
any orientation-independent interactions. We use a dispersion term
based on the SAFT-VR approach\cite{gil-villegas-1997-SAFT-VR}, which
has two free parameters (taken from Clark~\emph{et
  al}\cite{clark2006developing}): an interaction energy $\epsilondisp$
and a length scale $\lambdadisp R$.

@@
